Trends in Death License: Analyzing the Distribution of Death Licenses in Great Toronto Area

### 2024-09-27 starter_folder-main/paper/paper.qmd
let go

mydatapath = "~/Paper 1/starter_folder-main/data/analysis_data/analysis_data.csv"
data <- read_csv(mydatapath)
```

@@

The dataset was analyzed using R [@citeR] and downloaded using the R package opendatatoronto [@opendatatoronto]. Additionally, the packages tidyverse [@citetidyverse], lubridate [@citelubridate], readr [@citereadr], forecast [@citeforecast], ggplot [@citeggplot], dplyr [@citedplyr], here[@citehere], forecast [@citeforecast] have been used in data cleaning and visualization.

@@

# Display the first 10 rows of the data as a table
cleaned_data %>% 
  head(10) %>% 
  kable()

```
## Data Record 
```{r}
#| label: fig-01
#| fig-cap: "Histogram of Death Lincese From 2011 to 2024"
#| fig-align: 'center'
#| echo: false
#| warning: false
library(readr)

@@

data |>
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) |>
  group_by(year, place_of_death) |>
  summarise(death_licenses = sum(death_licenses), .groups = "drop") |>
  ggplot(mapping = aes(x = year, y = death_licenses, fill = place_of_death)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Year", y = "Number of Death Licenses",
    fill = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  scale_fill_brewer(palette = "Set1")

```

@@

# Process the data to get total death licenses per place of death
pie_data <- data %>%
  # Filter only the relevant locations
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>%
  # Group by place of death and sum the death licenses
  group_by(place_of_death) %>%
  summarise(death_licenses = sum(death_licenses, na.rm = TRUE)) %>%
  # Calculate percentage of death licenses
  mutate(percentage = death_licenses / sum(death_licenses) * 100,
         # Create label with specific percentage format
         label = paste0(round(percentage, 1), "%"))

# Create a pie chart with percentage labels
ggplot(pie_data, aes(x = "", y = percentage, fill = place_of_death)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y") + # Convert bar chart to pie chart
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) + # Add percentage labels
  labs(
    title = "Percentage of Death Licenses in Toronto and Outside City Limits in 2011-2024",
    fill = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  theme_void() + # Remove axis and grid lines for pie chart
  theme(legend.position = "right") # Position legend on the right

death_licenses_summary <- data %>%
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>%  # Filter for Toronto and Outside City Limits
  group_by(place_of_death, civic_centre) %>%
  summarise(total_death_licenses = sum(death_licenses, na.rm = TRUE)) %>%
  mutate(percentage = total_death_licenses / sum(total_death_licenses) * 100)  # Calculate percentage

```
In @fig-02, it is evident that Toronto accounts for 78.9% of the death licenses issued during this time period, while Outside City Limits comprises the remaining 21.1%. This significant disparity in population density is a consequence of urbanization, which results in diminished resource inflow to Outside City Limits.

@@

# Create the bar chart for the forecasted death licenses
ggplot(prediction_data, aes(x = month, y = death_licenses)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Predicted Number of Death Licenses",
    title = "Predicted Death Licenses for September to November 2024",
    caption = "Data source: Opendatatoronto."

  )
total_predicted_death_licenses <- sum(predicted_values)
# Print the total predicted death licenses

```

@@

# Create the line plot
ggplot(yearly_death_licenses, aes(x = year, y = total_death_licenses, color = place_of_death)) +
  geom_line(size = 1.2) +                                           # Line for each place of death
  geom_point(size = 3) +                                           # Points on the line for better visibility
  geom_hline(data = mean_death_licenses, aes(yintercept = mean_death_licenses, color = place_of_death), 
             linetype = "dashed", size = 1) +                    # Mean line
  theme_minimal() +
  labs(
    title = "Line Plot of Total Death License by Year",
    x = "Year",
    y = "Total Number of Death Licenses",
    caption = "Data source: Opendatatoronto.",
    color = "Place of Death"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better visibility

```
# Results {#sec-results}
## Death License with Month
The peak in death licenses consistently occurs in January each year. In recent years, extreme weather has been linked to an increase in mortality and may have an impact on the number of death licenses. As illustrated in @fig-05, the bar chart displays the total number of death licenses by month from 2011 to 2024 (excluding September to December 2024), revealing that January is the only month with a notable peak. In contrast, the remaining months generally fall within the mean line. According to data from the City of Toronto, there were a total of 262 extreme weather events from 2011 to 2022, with 115 occurring in January【@toronto2022extreme】. Given that Toronto's summer climate is more favorable compared to other regions, it may have a lower impact from extreme heat on mortality in summer. However, severe cold weather in winter could contribute to an increase in deaths, leading to a higher number of death licenses issued in January. A population-based study published in CMAJ, titled "Assessment of the Effect of Cold and Hot Temperatures on Mortality in Ontario, Canada," reveals that each 5°C change in daily temperature is estimated to induce 7 excess deaths per day during cold seasons and 4 excess deaths during warm seasons【@chen2016assessment】. This suggests that extreme winter weather may be a contributing factor to the increase in mortality which may have a positive relation with death license.
```{r}
#| label: fig-05
#| fig-cap: "Total Death License by Month"
#| fig-align: 'center'
#| echo: false
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(lubridate)

@@

# Create the bar chart with mean line
ggplot(monthly_death_licenses, aes(x = month, y = total_death_licenses)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Bar for each month
  geom_hline(yintercept = mean_death_licenses, color = "red", linetype = "dashed", size = 1) +  # Mean line
  theme_minimal() +
  labs(
    title = "Total Death Licenses by Month (2011-2024)",
    x = "Month",
    y = "Total Number of Death Licenses",
    caption = "Data source: Opendatatoronto."
  ) +
  scale_x_discrete(limits = month.abb) +  # Ensure months are in order
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better visibility

```

@@

# Process the data
plot_data <- data %>%
  # Filter only the relevant locations
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>%
  # Group by year, month, and place of death, and sum the death licenses
  group_by(year, month, place_of_death) %>%
  summarise(death_licenses = sum(death_licenses, na.rm = TRUE), .groups = 'drop') %>%
  # Create a new date column for easier plotting
  mutate(date = make_date(year, month, 1)) %>%
  # Filter to restrict time from 2015 to 2024
  filter(year >= 2016 & year <= 2023)

# Create a dot plot with clearer x-axis
ggplot(plot_data, aes(x = date, y = death_licenses, color = place_of_death)) +
  geom_point(size = 3) + # Use points for the dot plot
  scale_x_date(date_labels = "%b %Y", date_breaks = "5 months") + # Show every 4 months
  labs(
    title = "Monthly Death Licenses in Toronto and Outside City \n Limits (2015-2024)",
    x = "Year-Month",
    y = "Number of Death Licenses",
    color = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  theme_minimal() + # Clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis text for readability

```
# Discussion {#sec-discussion}

@@

Statement on LLM usage:LLMs were used in the making of this paper.

>>>>

mydatapath <- "~/Paper 1/starter_folder-main/data/analysis_data/analysis_data.csv"
data <- read_csv(mydatapath)
```

@@

The dataset was analyzed using R [@citeR] and downloaded using the R package opendatatoronto [@opendatatoronto]. Additionally, the packages tidyverse [@citetidyverse], lubridate [@citelubridate], readr [@citereadr], forecast [@citeforecast], ggplot [@citeggplot], dplyr [@citedplyr], here[@citehere], forecast [@citeforecast], styler[@citestyler] have been used in data cleaning and visualization.

@@

# Display the first 10 rows of the data as a table
cleaned_data %>%
  head(10) %>%
  kable()
```
## Data Record 
```{r}
#| label: fig-01
#| fig-cap: "Histogram of Death Lincese From 2011 to 2024"
#| fig-align: 'center'
#| echo: false
#| warning: false
library(readr)

@@

data |>
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) |>
  group_by(year, place_of_death) |>
  summarise(death_licenses = sum(death_licenses), .groups = "drop") |>
  ggplot(mapping = aes(x = year, y = death_licenses, fill = place_of_death)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Year", y = "Number of Death Licenses",
    fill = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  scale_fill_brewer(palette = "Set1")
```

@@

# Process the data to get total death licenses per place of death
pie_data <- data %>%
  # Filter only the relevant locations
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>%
  # Group by place of death and sum the death licenses
  group_by(place_of_death) %>%
  summarise(death_licenses = sum(death_licenses, na.rm = TRUE)) %>%
  # Calculate percentage of death licenses
  mutate(
    percentage = death_licenses / sum(death_licenses) * 100,
    # Create label with specific percentage format
    label = paste0(round(percentage, 1), "%")
  )

# Create a pie chart with percentage labels
ggplot(pie_data, aes(x = "", y = percentage, fill = place_of_death)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") + # Convert bar chart to pie chart
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) + # Add percentage labels
  labs(
    title = "Percentage of Death Licenses in Toronto and Outside City Limits in 2011-2024",
    fill = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  theme_void() + # Remove axis and grid lines for pie chart
  theme(legend.position = "right") # Position on the right

death_licenses_summary <- data %>%
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>% # Filter for Toronto and Outside City Limits
  group_by(place_of_death, civic_centre) %>%
  summarise(total_death_licenses = sum(death_licenses, na.rm = TRUE)) %>%
  mutate(percentage = total_death_licenses / sum(total_death_licenses) * 100) # Calculate percentage
```
In @fig-02, it is evident that Toronto accounts for 78.9% of the death licenses issued during this time period, while Outside City Limits comprises the remaining 21.1%. This significant disparity in population density is a consequence of urbanization, which results in diminished resource inflow to Outside City Limits.

@@

# Create the bar chart for the forecasted death licenses
ggplot(prediction_data, aes(x = month, y = death_licenses)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(
    x = "Month",
    y = "Predicted Number of Death Licenses",
    title = "Predicted Death Licenses for September to November 2024",
    caption = "Data source: Opendatatoronto."
  )
total_predicted_death_licenses <- sum(predicted_values)
# Print the total predicted death licenses
```

@@

# Create the line plot
ggplot(yearly_death_licenses, aes(x = year, y = total_death_licenses, color = place_of_death)) +
  geom_line(size = 1.2) + # Line for each place of death
  geom_point(size = 3) + # Points on the line for better visibility
  geom_hline(
    data = mean_death_licenses, aes(yintercept = mean_death_licenses, color = place_of_death),
    linetype = "dashed", size = 1
  ) + # Mean line
  theme_minimal() +
  labs(
    title = "Line Plot of Total Death License by Year",
    x = "Year",
    y = "Total Number of Death Licenses",
    caption = "Data source: Opendatatoronto.",
    color = "Place of Death"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better visibility
```
# Results {#sec-results}
## Death License with Month
The peak in death licenses consistently occurs in January each year. In recent years, extreme weather has been linked to an increase in mortality and may have an impact on the number of death licenses. As illustrated in @fig-05, the bar chart displays the total number of death licenses by month from 2011 to 2024 (excluding September to December 2024), revealing that January is the only month with a notable peak. In contrast, the remaining months generally fall within the mean line. According to data from the City of Toronto, there were a total of 262 extreme weather events from 2011 to 2022, with 115 occurring in January【@toronto2022extreme】. Given that Toronto's summer climate is more favorable compared to other regions, it may have a lower impact from extreme heat on mortality in summer. However, severe cold weather in winter could contribute to an increase in deaths, leading to a higher number of death licenses issued in January. A population-based study published in CMAJ, titled "Assessment of the Effect of Cold and Hot Temperatures on Mortality in Ontario, Canada," reveals that each 5°C change in daily temperature is estimated to induce 7 excess deaths per day during cold seasons and 4 excess deaths during warm seasons【@chen2016assessment】. This suggests that extreme winter weather may be a contributing factor to the increase in mortality which may have a positive relation with death license.
```{r}
#| label: fig-05
#| fig-cap: "Total Death License by Month"
#| fig-align: 'center'
#| echo: false
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(lubridate)

@@

# Create the bar chart with mean line
ggplot(monthly_death_licenses, aes(x = month, y = total_death_licenses)) +
  geom_bar(stat = "identity", fill = "skyblue") + # Bar for each month
  geom_hline(yintercept = mean_death_licenses, color = "red", linetype = "dashed", size = 1) + # Mean line
  theme_minimal() +
  labs(
    title = "Total Death Licenses by Month (2011-2024)",
    x = "Month",
    y = "Total Number of Death Licenses",
    caption = "Data source: Opendatatoronto."
  ) +
  scale_x_discrete(limits = month.abb) + # Ensure months are in order
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better visibility
```

@@

# Process the data
plot_data <- data %>%
  # Filter only the relevant locations
  filter(place_of_death %in% c("Toronto", "Outside City Limits")) %>%
  # Group by year, month, and place of death, and sum the death licenses
  group_by(year, month, place_of_death) %>%
  summarise(death_licenses = sum(death_licenses, na.rm = TRUE), .groups = "drop") %>%
  # Create a new date column for easier plotting
  mutate(date = make_date(year, month, 1)) %>%
  # Filter to restrict time from 2015 to 2024
  filter(year >= 2016 & year <= 2023)

# Create a dot plot with clearer x-axis
ggplot(plot_data, aes(x = date, y = death_licenses, color = place_of_death)) +
  geom_point(size = 3) + # Use points for the dot plot
  scale_x_date(date_labels = "%b %Y", date_breaks = "5 months") + # Show every 4 months
  labs(
    title = "Monthly Death Licenses in Toronto and Outside City \n Limits (2015-2024)",
    x = "Year-Month",
    y = "Number of Death Licenses",
    color = "Place of Death",
    caption = "Data source: Opendatatoronto."
  ) +
  theme_minimal() + # Clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis text for readability
```
# Discussion {#sec-discussion}

@@

Statement on LLM usage: LLMs were used in the making of this paper.
